---
title: "dgeAnalysis - EdgeR"
author:
- "Developed by LUMC-SASC team:"
- "- Tom Kuipers"
- "- Leon Mei"
- "- Davy Cats"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: paper
always_allow_html: true
params:
  data_samples:     ""
  data_counts:      ""
  data_annotation:  ""
  setGeneName:      ""
  cpm_value:        ""
  excluded_samples: ""
  design_base:      ""
  design_value:     ""
  matrix_v1:        ""
  matrix_v2:        ""
  alpha:            ""
---

# Reading packages and variables
Required packages are loaded and the input variables are set. The used input settings are shown.

```{r setup}

if (!require("knitr")) install.packages("knitr")
if (!require("SummarizedExperiment")) install.packages("SummarizedExperiment")
if (!require("edgeR")) install.packages("edgeR")
if (!require("tidyr")) install.packages("tidyr")
if (!require("scales")) install.packages("scales")
if (!require("plotly")) install.packages("plotly")

```


```{r variables}
# IMPORT VARIABLES

data_samples <- params$data_samples
data_counts <- params$data_counts
data_annotation <- params$data_annotation
setGeneName <- params$setGeneName
cpm_value <- params$cpm_value
excluded_samples <- params$excluded_samples
design_base <- params$design_base
design_value <- params$design_value
matrix_v1 <- params$matrix_v1
matrix_v2 <- params$matrix_v2
alpha <- params$alpha

```


```{r filterData}
# FILTER DATA IF NESSECARY

#Filter if nessecary
data_samples <- data_samples[!rownames(data_samples) %in% excluded_samples, , drop=FALSE]
data_counts <- data_counts[,!colnames(data_counts) %in% excluded_samples]

```

## Input values

```{r showVariables}
# SHOW VALUES OF VARIABLES

cpm_value
excluded_samples
alpha
design_base
design_value
matrix_v1
matrix_v2

```

# Process raw data
The raw data is processed and pre-filtered, to remove non feature counts. The data is placed in a format (DGE list) which can be used for later analsis.

```{r showRawData}
# SHOW RAW DATA

se <- readCountsFromTable(data_counts, data_samples)

alignmentSummaryPlot(se)
complexityPlot(se)

```


```{r filesIntoSE}
# READ ALL FILES INTO SE

se <- readCountsFromTable(data_counts[!grepl('^__', rownames(data_counts)),], data_samples)
se <- addSamplesFromTableToSE(se, data_samples)
if (!is.null(data_annotation)) {
  se <- addAnnotationsFromTableToSE(se, data_annotation)
}

```

## Create DGE list

```{r dge}
# GET AND CREATE DGE

dge <- DGEList(counts = assay(se), samples = colData(se), genes = rowData(se))
row.names(dge$genes) <- row.names(dge$counts)

dge <- dge[ rowSums( abs( dge$counts ) ) > 1, ]

tempDge <- dge
tempDge$counts <- cpm(dge, log = TRUE)
countDistributionLinePlot(tempDge)

```

# Filter raw data
The raw data is filtered based on the input values provided.

```{r selectedFeatures}
# GET SELECTED FEATURES

edger <- calcNormFactors( dge, method = "TMM")
counts <- cpm(edger, log = TRUE)
selectedFeatures <- rownames( edger )[ apply( counts, 1, function( v ) sum( v >= cpm_value ) ) >= 1/4 * ncol( counts ) ]

```


```{r highExpressed}
# GET HIGH EXPRESSED FEATURES

highExprDge <- dge[ selectedFeatures,, keep.lib.sizes = FALSE ]

```

# Normalization
The filtered data is normalized using TMM.

```{r normalize}
# NORMALIZE DATA

normDge <- calcNormFactors( highExprDge, method = "TMM")

tempDge <- normDge
tempDge$counts <- cpm(normDge, log = TRUE)
countDistributionLinePlot(tempDge)
samplePca2dPlot(tempDge, design_base, "PC1", "PC2")

```

# Create design and contrasts
The design matrix and contrasts are created, based on input values.

```{r design}
# CREATE DESIGN

get_design <- createDesign(normDge$samples, design_base, design_value, matrix_v1, matrix_v2)
get_design
normDge <- relevelSamples(normDge, design_base, design_value, matrix_v1, matrix_v2)
design <- model.matrix(eval(parse(text=get_design)), normDge$samples)
design

get_matrix1 <- createMatrix(normDge, design_base, design_value, matrix_v1)
get_matrix2 <- createMatrix(normDge, design_base, design_value, matrix_v2)
get_matrix1
get_matrix2

contrast <- createContrast(design, get_matrix1, get_matrix2)
contrast

```

# Perform analysis
The analysis is performed using edgeR with the normalized counts, design matrix and the provided contrasts.

```{r analysis}
# PERFORM ANALYSIS

normDge <- estimateDisp(normDge, design)
edfit <- glmFit(normDge, design)
efit <- glmLRT(edfit, contrast = contrast)

efit$DE <- decideTests(efit, p.value = alpha)
summary(efit$DE)

```

## Format DE output table

```{r create_deTab}
# CREATE deTab TABLE

deTab <- topTags(efit, n=Inf)$table
deTab <- deTab[order(rownames(deTab)),]
deTab$DE <- c(efit$DE[order(rownames(efit$DE)),])
deTab$adj.P.Val <- p.adjust(deTab$PValue, method = "BH")

deTab <- rename(deTab, "avgLog2CPM" = "logCPM")
deTab <- rename(deTab, "avgLog2FC" = "logFC")
deTab <- rename(deTab, "P.Value" = "PValue")

```


```{r changeToSymbol}
# CHANGE GENE ID TO SYMBOL IF NESSECARY

if (setGeneName == "symbol" && !is.null(data_annotation)) {
  tempCol <- rownames(deTab)
  rownames(deTab) <- make.names(deTab$geneName, unique=TRUE)
  deTab$geneName <- tempCol
  colnames(deTab)[1] <- "geneId"
  rownames(normDge$counts) <- rownames(deTab)
}

```

## DE results

```{r showDGEResult}
# SHOW DGE RESULTS

ma_plot(deTab)
pValuePlot(deTab)

```

# Save results
Results are saved, so they can be retrieved by the application.

```{r save}
# SAVE ANALYSIS

normDge$counts <- cpm(normDge, log = TRUE)
save(deTab, normDge, file="analysis.RData")

```

# Session info
Showing all used packages including version numbers.

```{r session}
# INFO

sessionInfo()

```
