---
title: "dgeAnalysis - DESeq2"
author:
- "Developed by LUMC-SASC team:"
- "- Tom Kuipers"
- "- Leon Mei"
- "- Davy Cats"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: paper
always_allow_html: true
params:
  data_samples:     ""
  data_counts:      ""
  data_annotation:  ""
  setGeneName:      ""
  cpm_value:        ""
  excluded_samples: ""
  design_base:      ""
  design_value:     ""
  matrix_v1:        ""
  matrix_v2:        ""
  alpha:            ""
---

# Reading packages and variables
Required packages are loaded and the input variables are set. The used input settings are shown.

```{r setup}

if (!require("knitr")) install.packages("knitr")
if (!require("SummarizedExperiment")) install.packages("SummarizedExperiment")
if (!require("DESeq2")) install.packages("DESeq2")
if (!require("tidyr")) install.packages("tidyr")
if (!require("scales")) install.packages("scales")
if (!require("plotly")) install.packages("plotly")

```


```{r variables}
# IMPORT VARIABLES

data_samples <- params$data_samples
data_counts <- params$data_counts
data_annotation <- params$data_annotation
setGeneName <- params$setGeneName
cpm_value <- params$cpm_value
excluded_samples <- params$excluded_samples
design_base <- params$design_base
design_value <- params$design_value
matrix_v1 <- params$matrix_v1
matrix_v2 <- params$matrix_v2
alpha <- params$alpha

```


```{r filterData}
# FILTER DATA IF NECESSARY

data_samples <- data_samples[!rownames(data_samples) %in% excluded_samples, , drop=FALSE]
data_counts <- data_counts[,!colnames(data_counts) %in% excluded_samples]

```

## Input values

```{r showVariables}
# SHOW VALUES OF VARIABLES

cpm_value
excluded_samples
alpha
design_base
design_value
matrix_v1
matrix_v2

```

# Process raw data
The raw data is processed and pre-filtered, to remove non feature counts. The data is placed in a format (DGE list) which can be used for later analsis.

```{r showRawData}
# SHOW RAW DATA

se <- readCountsFromTable(data_counts, data_samples)
se <- addSamplesFromTableToSE(se, data_samples)

alignmentSummaryPlot(se)
complexityPlot(se)

```

## Create DGE list

```{r filesIntoSE}
# READ ALL FILES INTO SE

se <- readCountsFromTable(data_counts[!grepl('^__', rownames(data_counts)), ], data_samples)
se <- addSamplesFromTableToSE(se, data_samples)
if (!is.null(data_annotation)) {
  se <- addAnnotationsFromTableToSE(se, data_annotation)
}

dge <- DGEList(counts = assay(se),
               samples = colData(se),
               genes = rowData(se))
dge <- dge[rowSums(abs(dge$counts)) > 1,]

tempDge <- dge
tempDge$counts <- cpm(dge, log = TRUE, prior.count = 1)
countDistributionLinePlot(tempDge)

dge <- DGEList(counts = assay(se), samples = colData(se))

```

# Create design and prepare analysis
The design matrix is created and the raw data is filtered based on the input values provided.

```{r selectedFeatures}
# GET SELECTED FEATURES

get_design <- createDesign(dge$samples, design_base, design_value, matrix_v1, matrix_v2)
get_design
dge <- relevelSamples(dge, design_base, design_value, matrix_v1, matrix_v2)
design <- model.matrix(eval(parse(text = get_design)), dge$samples)
design

analysis <- DESeqDataSet(se, design = design)
analysis <- analysis[rowSums(abs(assay(analysis))) > 1,]

# FILTER IF NESSECARY
counts <- cpm(counts(analysis), log = TRUE, prior.count = 1)
selectedFeatures <- rownames(analysis)[apply(counts, 1, function(v)
  sum(v >= cpm_value)) >= 1 / 4 * ncol(counts)]
analysis <- analysis[selectedFeatures, ]

```

# Perform analysis
The analysis is performed with the filtered data using DESeq2.

```{r startAnalysis}
# START ANALYSIS

analysis <- DESeq(analysis)

```

# Normalization
Extract the normalized data from the analysis results.

```{r getNormalizedCounts}
# GET NORMALIZED COUNTS

getSize <- estimateSizeFactors(analysis)
normDge <- DGEList(counts = data.frame(counts(getSize, normalized = TRUE)), samples = dge$samples)
normDge$counts <- cpm(normDge, log = TRUE, prior.count = 1)

countDistributionLinePlot(normDge)
pcaPlot(normDge, design_base, "PC1", "PC2")

```

# Collect DE results
Create the contrast matrix and collect DE results based on this matrix.

```{r getResults}
# GET RESULTS

get_matrix1 <- createMatrix(normDge, design_base, design_value, matrix_v1)
get_matrix2 <- createMatrix(normDge, design_base, design_value, matrix_v2)
get_matrix1
get_matrix2

contrast <- createContrast(design, get_matrix1, get_matrix2)
contrast

deTab <- results(analysis, contrast = contrast, alpha = alpha)
summary(deTab)

```

## Format DE output table

```{r create_deTab}
# CREATE deTab TABLE

deTab <- data.frame(deTab[, c("log2FoldChange", "pvalue", "padj")])
colnames(deTab) <- c("avgLog2FC", "P.Value", "adj.P.Val")
deTab$avgLog2CPM <- rowMeans(normDge$counts)

# SET DE GENES
deTab$DE <- 0
deTab$DE[deTab$adj.P.Val < alpha & deTab$avgLog2FC < 0] <- -1
deTab$DE[deTab$adj.P.Val < alpha & deTab$avgLog2FC > 0] <- 1

```

## Add annotation to DE table

```{r addAnnotation}
# ADD ANNOTATION TO deTab

if (!is.null(data_annotation)) {
  deTab <- merge(data_annotation, deTab, by = 0, all = FALSE)
  rownames(deTab) <- deTab$Row.names
  deTab$Row.names <- NULL
}

#ORDER deTab TABLE
deTab <- rename(deTab, "FDR" = "adj.P.Val")
deOrder <- c("avgLog2CPM", "avgLog2FC", "P.Value", "FDR", "DE")

if (!is.null(data_annotation)) {
  deTab <- deTab[colnames(deTab)
                 [c(1, match(deOrder, names(deTab)), 2:(ncol(deTab) - 5))]]
} else {
  deTab <- deTab[deOrder]
}

```


```{r changeToSymbol}
# CHANGE GENE ID TO SYMBOL IF NECESSARY

if (setGeneName == "symbol" && !is.null(data_annotation)) {
  tempCol <- rownames(deTab)
  rownames(deTab) <- make.names(deTab$geneName, unique = TRUE)
  deTab$geneName <- tempCol
  colnames(deTab)[1] <- "geneId"
  rownames(normDge$counts) <- rownames(deTab)
}

```

## DE results

```{r showDGEResult}
# SHOW DGE RESULTS

ma_plot(deTab)
pValuePlot(deTab)

```

# Save results
Results are saved, so they can be retrieved by the application.

```{r save}
# SAVE ANALYSIS

deTab <- deTab[order(deTab$FDR),]
save(deTab, normDge, file = "analysis.RData")

```

# Session info
Showing all used packages including version numbers.

```{r session}
#INFO

sessionInfo()

```
