---
title: "dgeAnalysis - gProfiler2 Enrichment"
author:
- "Developed by LUMC-SASC team:"
- "- Tom Kuipers"
- "- Leon Mei"
- "- Davy Cats"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: paper
always_allow_html: true
params:
  geneList:     ""
  organism:     ""
  significant:  ""
  alfa:         ""
  database:     ""
---

# Reading packages and variables
Required packages are loaded and the input variables are set. The used input settings are shown.

```{r setup}

if (!require("gprofiler2")) install.packages("gprofiler2")

```


```{r variables}
# IMPORT VARIABLES

geneList <- strsplit(params$geneList,"\n")[[1]]
organism <- params$organism
significant <- eval(parse(text=params$significant))
alfa <- params$alfa
database <- params$database

```

## Input values

```{r showVariables}
# SHOW VALUES OF VARIABLES

length(geneList)
organism
significant
alfa
database

```

## Run enrichment
Enrichment analysis is performed using gProfiler2.

```{r runEnrichment}
# SHOW VALUES OF VARIABLES

enrich <- gost(
  query = geneList,
  organism = organism,
  ordered_query = FALSE,
  significant = significant,
  user_threshold = alfa,
  correction_method = "bonferroni",
  domain_scope = "annotated",
  sources = database,
  evcodes = TRUE
)

```

# Show results
Show gProfiler enrichment plot

```{r showResult}
# Show results

enrich$result$query <- ""
gostplot(enrich, capped = FALSE, interactive = TRUE)

```


# Save results
Results are saved, so they can be retrieved by the application.

```{r save}
# SAVE ANALYSIS

save(enrich, file = "enrichment.RData")

```

# Session info
Showing all used packages including version numbers.

```{r session}
# INFO

sessionInfo()

```
